MyPDDL: Tools for efficiently creating PDDL domains and problems

### 2019-07-03 paper.tex
upd

\usepackage{hyperref}
\usepackage{graphicx}
\graphicspath{{img/}}
\usepackage{tabularx}
\newcommand{\tableheadline}[1]{\multicolumn{1}{c}{#1}}
\usepackage{dirtree}
\usepackage{rotating}
\usepackage{caption}
\usepackage{booktabs}
% More space between rows:		
\renewcommand{\arraystretch}{1.1}		
% Slightly more space between columns:		
\setlength{\tabcolsep}{8pt}
\usepackage{minted}
\definecolor{dhscodebg}{rgb}{0.95,0.95,0.95} % listing background

\usepackage{xspace}
\renewcommand\UrlFont{\color{blue}\rmfamily}

\newcommand{\mypddl}{\textsc{myPddl}\xspace}
\newcommand{\mypddlclojure}{\textsc{myPddl-clojure}\xspace}
\newcommand{\mypddlsyntax}{\textsc{myPddl-syntax}\xspace}
\newcommand{\mypddldiagram}{\textsc{myPddl-diagram}\xspace}
\newcommand{\mypddlnew}{\textsc{myPddl-new}\xspace}
\newcommand{\mypddlide}{\textsc{myPddl-ide}\xspace}
\newcommand{\mypddlsnippet}{\textsc{myPddl-snippet}\xspace}
\newcommand{\mypddldistance}{\textsc{myPddl-distance}\xspace}
\newcommand{\mypddlplan}{\textsc{myPddl-plan}\xspace}
\newcommand{\pddlstudio}{\textsc{pddl studio}\xspace}
\newcommand{\itsimple}{\textsc{itSimple}\xspace}
\newcommand{\pddlmode}{\textsc{pddl}-mode\xspace}
\newcommand{\pddl}{\textsc{pddl}\xspace}
\newcommand{\uml}{\textsc{uml}\xspace}
\newcommand{\ide}{\textsc{ide}\xspace}
\newcommand{\sublimetext}{Sublime Text\xspace}

@@

\keywords{PDDL \and Planning}
\end{abstract}
%
%

@@

Being a key aspect of artificial intelligence, \emph{planning} is
concerned with devising a sequence of actions to achieve a desired
goal \cite{helmert2008understanding}. It can be both a tool to create
automated systems and a means to support and understand human behavior
\cite{konar1999artificial}. However, the effectiveness of planning
largely depends on the quality of the problem formalization
\cite{shah2013knowledge,keps2014}. To ensure a standardized modeling
format, the Planning Domain Definition Language (\textsc{pddl})
\cite{mcdermott1998pddl} was developed and has become the de facto
standard for the description of planning tasks
\cite{ilghami2005extension}. The discipline that deals with the
integration of world information into a computer system via a human
expert is called knowledge engineering \cite{feigenbaum1983fifth}.
While automated planning could save vast amounts of time if everything
works as intended, creating the planning task specifications is a
complex task that can be error-prone and cumbersome.

The specification of Artificial Intelligence (AI) planning can be
time-consuming and cubersome. While there is a large spectrum of
different AI languages, the de facto standard language is PDDL
(\emph{Planning Domain Definition Language}). However, so far its use
is predominantly confined to academic labs. One of the reasons seems
to be the missing support of knowledge engieering tools so simplify
the task of specifying complex planning domains.

In this chapter, we describe \mypddl
(Figure~\ref{fig:mypddl-overview}), a knowledge engineering toolkit
that supports the AI knowledge engineer in the development of complex
AI planning problems. \mypddl is intended to support knowledge
engineers throughout the entire design cycle of specifying planning
tasks. In the initial stages, it allows for the creation of structured
\textsc{pddl} projects that should encourage a disciplined design
process. With the help of snippets, i.e. code templates, often used
constructs can be inserted in \textsc{pddl} files. A syntax
highlighting feature that speeds up the error-detection can come in
handy in intermediate stages. Understanding the textual representation
of complex type hierarchies in domain files can be confusing, so an
additional tool enables their visualization. \textsc{pddl}'s limited
modeling capabilities were bypassed by developing an interface that
converts \pddl code into Clojure \cite{hickey2008clojure} code and
vice versa.  Within this project, the interface was employed for a
feature that calculates distances between objects specified in a
problem model, but the interface provides numerous other possibilities
and could also be used to further automate the modeling process. All
of the features were integrated into the customizable and extensible
Sublime Text \cite{sublimetext3} editor. Since the main aim in the
development of the toolkit was for it to be easy to use and maintain,
it is evaluated with regard to these criteria. The usability was
assessed by means of a user test with eight subjects that had no prior
experience with artificial intelligence planning. The results indicate
that both error-detection and the understanding of a given domain can
be facilitated by \mypddl.

\begin{figure}[h]
  \centering
  \begin{tikzpicture}[
  level 1/.style={sibling distance=40mm},
  edge from parent/.style={->,draw},
  >=latex]

  % root of the the initial tree, level 1
  \node[] (ide) {-IDE};
\node[below of = ide,root] {myPDDL}
% The first level, as children of the initial tree
  child {node[level 2] (c1) {-new}}
  child {node[level 2] (c2) {-syntax}}
  child {node[level 2] (c3) {-snippet}}
  child {node[level 2] (c4) {-distance}}
  child {node[level 2] (c5) {-diagram}};

% The second level, relatively positioned nodes
\begin{scope}[every node/.style={level 3}]
\node [below of = c1, xshift=15pt] (c11) {Setting shape};
\node [below of = c11] (c12) {Choosing color};
\node [below of = c12] (c13) {Adding shading};

\node [below of = c2, xshift=15pt] (c21) {Using a Matrix};
\node [below of = c21] (c22) {Relatively};
\node [below of = c22] (c23) {Absolutely};
\node [below of = c23] (c24) {Using overlays};

\node [below of = c3, xshift=15pt] (c31) {Default arrows};
\node [below of = c31] (c32) {Arrow library};
\node [below of = c32] (c33) {Resizing tips};
\node [below of = c33] (c34) {Shortening};
\node [below of = c34] (c35) {Bending};
\end{scope}

% lines from each level 1 node to every one of its "children"
\foreach \value in {1,2,3}
  \draw[->] (c1.195) |- (c1\value.west);

\foreach \value in {1,...,4}
  \draw[->] (c2.195) |- (c2\value.west);

\foreach \value in {1,...,5}
  \draw[->] (c3.195) |- (c3\value.west);
\end{tikzpicture}
  \caption{\mypddl is a highly customizable and extensible modular
    system, designed for supporting knowledge engineers in the process
    of writing, analyzing and expanding PDDL files and thereby
    promoting the collaboration between knowledge engineers and the
    use of PDDL in real-world applications. It consists of the parts
    shown in the figure.}
  \label{fig:mypddl-overview}
\end{figure}

The remainder of this chapter is structured as
follows. Section~\ref{sec:related-work} compares other PDDL editors
and \pddl knowledge engineering tools. Section~\ref{sec:myPDDL}
describes the different modules of \mypddl and their design
principles. Section~\ref{sec:valid-eval} compares \mypddl to the other
existing tools using a benchmark
validation. Section~\ref{sec:conclusion} concludes the chapter and
outlines future work.

@@

This section introduces knowledge engineering tools that allow editing
\textsc{pddl} files in a textual environment. After introducing the
tools, they are compared and their shortcomings are discussed to set
the stage for \mypddl.

\pddlstudio \cite{plch2012inspect} is an application for creating and
managing \pddl projects, i.e. a collection of \pddl
files. \pddlstudio 's integrated development environment (\ide) was
inspired by Microsoft Visual Studio and imperative programming
paradigms. Its main features are syntax highlighting, error detection,
context sensitive code completion, code folding, project management,
and planner integration. \pddlstudio 's error detection can recognize
both syntactic (missing keywords, parentheses, etc.) and semantic
(wrong type of predicate parameters, misspelled predicates, etc.)
errors.

A major drawback of \pddlstudio is that it is not updated regularly
and only supports \pddl 1.2. Later \pddl versions contain several
additional features such as durative actions, numeric fluents, and
plan metrics \cite{edelkamp2004pddl2}.

\itsimple \cite{vaquero2005itsimple} follows a graphical approach
using Unified Modeling Language (\uml) diagrams. In the process
leading up to \itsimple, \textsc{uml.p} (\uml in a Planning Approach)
was proposed, a \uml variant specifically designed for modeling
planning domains and problems \cite{vaquero2006use}.

\itsimple's modeling workflow is unidirectional as changes in the
\pddl domain do not affect the \uml model and \uml models have to be
modeled manually, meaning that they cannot by generated from
\pddl. However, \cite{tonidandel2006reading} present a translation
process from a \pddl domain specification to an object-oriented
\textsc{uml.p} model as a possible integration for \itsimple. This
translation process makes extensive semantic assumptions for \pddl
descriptions. For example, the first parameter in the
\texttt{:parameters} section of an action is automatically declared as
a subclass of the default class \texttt{Agent}, and the method is
limited to predicates with a maximum arity of two. The current
version of \itsimple does not include the translation process from
\pddl to \uml.

Starting in version 4.0, \itsimple expanded its features to allow the
creation of \pddl projects from scratch (i.e. without the \uml to
\pddl translation process) \cite{vaquero2012itsimple4}. Thus far, the
\pddl editing features are basic. A minimal syntax highlighting
feature recognizes \pddl keywords, variables, and comments. \itsimple
also provides templates for \pddl constructs, such as requirement
specifications, predicates, actions, initial state, and goal
definitions.

Both \pddlstudio and \itsimple do not build on existing editors and
therefore cannot fall back on refined implementations of features that
have been modified and improved many times throughout their existence.

\pddlmode\footnote{http://rakaposhi.eas.asu.edu/planning-
  list-mailarchive/msg00085.html } for the widely used Emacs editor
builds on the sophisticated features of Emacs and uses its
extensibility and customizability. It provides syntax highlighting by
way of basic pattern matching of keywords, variables, and
comments. Additional features are automatic indentation and code
completion as well as bracket matching. Code snippets for the creation
of domains, problems, and actions are also available. Finally,
\textsc{pddl}-mode keeps track of action and problem declarations by
adding them to a menu and thus intending to allow for easy and fast
code navigation.

@@

The tool editor.planning.domains allows for editing \textsc{pddl}
files online. Its feature comprise syntax highlighting, code folding,
\textsc{pddl}-specific auto-completion, and multi-tab support. The
editor is part of the Planning.Domains initiative which aims at
providing three pillars to the planning community: (1) an API for
existing planning problems; (2) a planner-in-the-cloud service; and
(3) an online PDDL editor. In contrast to \mypddl,
editor.planning.domains does not support context-aware syntax
highlighting: for example, editor.planning.domains highlights key
words even if they are misplaced. In contrast, \mypddl is aware of the
scope of a construct and highlights key words accordingly.

The tool
vscode-pddl\footnote{\url{https://marketplace.visualstudio.com/items?itemName=jan-dolejsi.pddl}}
offers a wide range of editing functions, such as syntax highlighting,
code completion, and code snippets. Since it is implemented as a
plugin for the editor Visual Studio Code, it can fall back on the the
extensibility and maturity of this code editor. In contrast to
\mypddl, it does not offer domain visualization nor an interface with
a programming language in order to extend \pddl's limited modeling
capacities.

In sum, there is currently no tool available supporting all features
of \pddl 3.1, nor all the steps in the modeling process.

\mypddl is a highly customizable and extensible modular system,
designed for supporting knowledge engineers in the process of writing,
analyzing and expanding \pddl files and thereby promoting the
collaboration between knowledge engineers and the use of \pddl in
real-world applications. \mypddl consists of the following integral
parts:

\begin{description}
\item[\mypddlide] is an integrated development environment (IDE) for
  the use of \mypddl in the text and code editor \emph{Sublime
    Text}\footnote{\url{http://www.sublimetext.com}}. Since
  \mypddlsnippet and \textsc{-syntax} are devised explicitly for
  \sublimetext, their integration is implicit. The other tools (-new,
  -diagram, -distance) can be used independently of \sublimetext with
  the command-line interface and any \pddl file, but were also
  integrated into the editor.

\item[\mypddlsyntax] is a context-aware syntax highlighting feature
  for \sublimetext. It distinguishes all \pddl constructs up to
  version 3.1. Using regular expressions that can recognize both the
  start and the end of code blocks by means of a sophisticated pattern
  matching heuristic, \mypddlsyntax identifies \pddl code blocks and
  constructs and divides them into so called scopes, i.e. named
  regions. \sublimetext colorizes the code elements via the assigned
  scope names and in accordance with the current color scheme. These
  scopes allow for a fragmentation of the \pddl files, so that
  constructs are only highlighted if they appear in the correct
  context. Thus missing brackets, misplaced expressions and misspelled
  keywords are visually distinct and can be identified (see
  Figure~\ref{fig:syntax}).

  \begin{figure}
    \centering
    \includegraphics[width=0.6\textwidth]{coffee-yes.pdf}
    \caption{Syntax highlighting using \mypddlide. White
      text contains errors.}
\label{fig:syntax}
  \end{figure}

\item[\mypddlnew] helps to organize \pddl projects by generating the
following folder structure:

\begin{figure}[h] 
  \dirtree{%
  .1 project-name/.
  .2 domains/.
  .2 problems/.
  .3 p01.pddl.
  .2 solutions/.
  .2 domain.pddl.
  .2 README.md.
  .2 plan.
  }
\end{figure}

The domain file \texttt{domain.pddl} and the problem file
\texttt{p01.pddl} initially contain corresponding \pddl skeletons
which can also be customized. All problem files that are associated
with one domain file are collected in the folder
\texttt{problems/}. \texttt{README.md} is a Markdown file, which is
intended for (but not limited to) information about the author(s) of
the project, contact information, informal domain and problem
specifications, and licensing information.  Markdown files can be
converted to \textsc{html} by various hosting services (like GitHub or
Bitbucket). The basic planner integration \mypddlplan provided by the
file \texttt{plan} is described below.

\item[\mypddlsnippet] provides code skeletons, i.e. templates for
  often used pddl constructs such as domains, problems, type and
  function declarations, and actions. They can be inserted by typing a
  triggering keyword. Almost all \pddl actions consist of these same parts. And this is only
an example. Knowledge engineers have to use the same constructs again
and again when writing and extending \pddl files. This is where code
snippets can come in handy. To facilitate and speed up the implementation of
standard constructs, \mypddl-snippet provides code skeletons, i.e.
templates for often used \pddl constructs. They can be inserted by
typing a triggering keyword (trigger). Table \ref{tab:snippets}
displays descriptions of all available snippets and the corresponding
trigger.

\begin{table}[htb]
\centering
\begin{tabular}{ll}
  \tableheadline{snippet description} & \tableheadline{trigger}                   \\
  \hline
  domain skeleton                     & \texttt{domain}                           \\
  problem skeleton                    & \texttt{problem}                          \\
  type declaration                    & \texttt{t1, t2, ...}                      \\
  typed predicate declaration         & \texttt{p1, p2, ...}                     \\ 
  typed function declaration          & \texttt{f1, f2, ...}                     \\
  action skeleton                     & \texttt{action}, \texttt{durative-action} \\
\end{tabular}\caption[Available snippets in \mypddl-snippet]{\label{tab:snippets}The snippets that can be inserted into \pddl files by typing the trigger.}

\end{table}

For example, typing \texttt{action} and pressing the tabulator key (this
function defaults to the tabulator key, but the key can be customized)
inserts the action skeleton in Listing \ref{ls:action-skeleton}. \pddl
constructs with a specified arity can be generated by adding the arity
number to the trigger (\texttt{p2} would insert the binary predicate
template \texttt{(pred-name~?x~-~object~?y~-~object)}.

Once the snippet has been inserted, skipping from blank to blank is
enabled by pressing the tabulator key. This allows for a fast
navigation within the snippet.

Every snippet is stored in a separate file, located in the folder
\texttt{Packages/PDDL/} of Sublime Text. New snippets can be added and
existing snippets can be customized (by changing the template or the
trigger) in this folder.

\item[\mypddlclojure] provides a preprocessor for \pddl files to
  bypass \pddl's limited mathematical capabilities, thus reducing
  modeling time without overcharging planning algorithms. Since \pddl
  is used to create more and more complex domains
  \cite{goldman2012type,guerin2012academic}, needing the square root
  function for a distance optimization problem or the logarithmic
  function for modeling an engineering problem does not seem to be a
  far-fetched scenario. However, \pddl's calculating capabilities are
  limited \cite{parkinson2012increasing}. While these mathematical
  operations are currently not supported by \pddl itself,
  preprocessing \pddl files in a programming language and then
  hardcoding the results back into the file seems to be a reasonable
  workaround. With the help of such an interface, the modeling time
  can reduced an even partly automated (see \ref{subsec:loc} the
  distance calculator \mypddl-distance).  We decided to use Clojure
  \cite{hickey2008clojure}, a modern Lisp dialect that runs on the
  Java Virtual Machine (\textsc{jvm}) \cite{lindholm2014java},
  facilitating input and output of the Lisp-style \pddl
  constructs. The interface is provided as a Clojure library and based
  on two methods:
\begin{description}
\item[{read-construct(keyword, file)}] Allows for the extraction of
  code blocks from \pddl files. Listing \ref{src:read-construct} shows
  an example in which the goal state of \emph{Gary's Huge Problem} is
  extracted.
\end{description}
\begin{description}
\item[{add-construct(file, block, part)}] Provides a means for adding
constructs to a specified code block in \pddl problem files. This
is illustrated in Listing \ref{ls:using-interface} and
\ref{ls:after-interface} where the predicate \texttt{(hungry gisela)} is
added to the \texttt{(:init~...)} block of \emph{Gary's Huge Problem}.
\end{description}

\begin{listing}[H]
\begin{minted}[fontsize=\small,bgcolor=dhscodebg,rulecolor=\color{gray!40},frame=lines,framesep=5\fboxrule,framerule=1pt,tabsize=2]{clojure}
(read-construct :goal "garys-huge-problem.pddl")
;;=> ((:goal (exploited magicfailureapp)))
\end{minted}
\caption[Extracting the goal state using the interface]{\label{src:read-construct}Extracting the goal state of \emph{Gary's Huge Problem} by using the interface.}
\end{listing}

\begin{description}
\item[{add-construct(file, block, part)}] Provides a means for adding
constructs to a specified code block in \pddl problem files. This
is illustrated in Listing \ref{ls:using-interface} and
\ref{ls:after-interface} where the predicate \texttt{(hungry gisela)} is
added to the \texttt{(:init~...)} block of \emph{Gary's Huge Problem}.
\end{description}

\begin{listing}[H]
\begin{minted}[fontsize=\small,bgcolor=dhscodebg,rulecolor=\color{gray!40},frame=lines,framesep=5\fboxrule,framerule=1pt,tabsize=2]{clojure}
(add-construct "garys-huge-problem.pddl" :init '((hungry gisela)))
\end{minted}
\caption[Adding a predicate using the
interface]{\label{ls:using-interface}Adding the predicate
  \texttt{(hungry~ gisela)} to the \texttt{(:init~...)} block of
  \emph{Gary's Huge Problem} using the PDDL/Clojure interface.}
\end{listing}

>>>>

\usepackage{hyperref}
\usepackage{graphicx}
\graphicspath{{img/}}
\usepackage{tabularx}
\newcommand{\tableheadline}[1]{\multicolumn{1}{c}{#1}}
\usepackage{dirtree}
\usepackage{rotating}
\usepackage{caption}
\usepackage{booktabs}
% More space between rows:		
\renewcommand{\arraystretch}{1.1}		
% Slightly more space between columns:		
\setlength{\tabcolsep}{8pt}
\usepackage{minted}
\definecolor{dhscodebg}{rgb}{0.95,0.95,0.95} % listing background
\DeclareUnicodeCharacter{2212}{-}

\usepackage{xspace}
\renewcommand\UrlFont{\color{blue}\rmfamily}
\usepackage{csquotes}

\newcommand{\mypddl}{\textsc{myPddl}\xspace}
\newcommand{\mypddlclojure}{\textsc{myPddl-clojure}\xspace}
\newcommand{\mypddlsyntax}{\textsc{myPddl-syntax}\xspace}
\newcommand{\mypddldiagram}{\textsc{myPddl-diagram}\xspace}
\newcommand{\mypddlnew}{\textsc{myPddl-new}\xspace}
\newcommand{\mypddlide}{\textsc{myPddl-ide}\xspace}
\newcommand{\mypddlsnippet}{\textsc{myPddl-snippet}\xspace}
\newcommand{\mypddldistance}{\textsc{myPddl-distance}\xspace}
\newcommand{\mypddlplan}{\textsc{myPddl-plan}\xspace}
\newcommand{\pddlstudio}{\textsc{pddl studio}\xspace}
\newcommand{\epd}{editor.planning.domains\xspace}
\newcommand{\itsimple}{\textsc{itSimple}\xspace}
\newcommand{\pddlmode}{\textsc{pddl}-mode\xspace}
\newcommand{\pddl}{\textsc{pddl}\xspace}
\newcommand{\uml}{\textsc{uml}\xspace}
\newcommand{\ide}{\textsc{ide}\xspace}
\newcommand{\sublimetext}{Sublime Text\xspace}

@@

\keywords{PDDL \and Planning \and Knowledge Engineering}
\end{abstract}

@@

Being a key aspect of artificial intelligence, \emph{planning} is
concerned with devising a sequence of actions to achieve a desired
goal \cite{helmert2008understanding}. It can be both a tool to create
automated systems and a means to support and understand human behavior
\cite{konar1999artificial}. However, the effectiveness of planning
largely depends on the quality of the problem formalization
\cite{shah2013knowledge,keps2014}. To ensure a standardized modeling
format, \pddl (\emph{Planning Domain Definition Language})
\cite{mcdermott1998pddl} was developed and has become the de facto
standard for the description of planning tasks
\cite{ilghami2005extension}. The discipline that deals with the
integration of world information into a computer system via a human
expert is called knowledge engineering \cite{feigenbaum1983fifth}.
While automated planning could save vast amounts of time if everything
works as intended, creating the planning task specifications is a
complex task that can be error-prone and cumbersome.

In this chapter, we describe \mypddl
(Figure~\ref{fig:mypddl-overview}), a knowledge engineering toolkit
that supports the knowledge engineer in the entire design cycle of
specifying planning tasks. In the initial stages, it allows for the
creation of structured \pddl projects that should encourage a
disciplined design process. With the help of snippets, that is, code
templates, often used syntactic constructs can be inserted into \pddl
files. A syntax highlighting feature that speeds up the
error-detection can come in handy in intermediate
stages. Understanding the textual representation of complex type
hierarchies in domain files can be confusing, so an additional tool
enables their visualization. \pddl's limited modeling capabilities
were bypassed by developing an interface that converts \pddl code into
code of the functional programming language Clojure
\cite{hickey2008clojure} and vice versa.  Within this project, the
interface was employed for a feature that calculates distances between
objects specified in a problem model, but the interface provides
numerous other possibilities and could also be used to further
automate the modeling process. All of the features were integrated
into the customizable and extensible Sublime Text \cite{sublimetext3}
editor. Since the main aim in the development of the toolkit was for
it to be easy to use and maintain, it is evaluated with regard to
these criteria. The usability was assessed by means of a user test
with eight subjects that had no prior experience with AI planning. The
results indicate that both error-detection and the understanding of a
given domain can be facilitated by \mypddl. This chapter is an
extended version of work published in a previous paper~\cite{StrKir2014:aai}.

\begin{figure}[H]
  \centering
  \tikzstyle{every node}=[draw=black,anchor=west,
  align=left,
  scale=0.7]
  \begin{tikzpicture}[%
  grow via three points={one child at (0.5,-0.7) and
    two children at (0.5,-0.7) and (0.5,-1.4)},
  edge from parent path={(\tikzparentnode.south) |- (\tikzchildnode.west)}]
  \node {~~~\textbf{myPDDL}~~~}
    child { node {\textbf{-new}\\Create new \pddl project}}		
    child { node {\textbf{-syntax}\\Context-aware syntax highlighting}}
    child { node {\textbf{-snippet}\\Code templates}}
    child { node {\textbf{-clojure}\\Preprocessor for \pddl files via Clojure}}
    child { node {\textbf{-distance}\\Distance calculation in spatial domains}}
    child { node {\textbf{-diagram}\\Visualization of type hierarchy}}
    child { node {\textbf{-plan}\\Planner integration}};
  \end{tikzpicture}
    \caption{\mypddl is a highly customizable and extensible modular
    system, designed for supporting knowledge engineers in the process
    of writing, analyzing and expanding PDDL files and thereby
    promoting the collaboration between knowledge engineers and the
    use of PDDL in real-world applications. It consists of the parts
    shown in the figure.}
  \label{fig:mypddl-overview}  
\end{figure}

The remainder of this chapter is structured as
follows. Section~\ref{sec:related-work} compares \pddl knowledge
engineering tools to lay a foundation for
\mypddl. Section~\ref{sec:mypddl} describes the different modules of
\mypddl and their design principles. Section~\ref{sec:valid-eval}
evaluates \mypddl via a user test. Section~\ref{sec:conclusion}
concludes the chapter and outlines future work.

@@

This section introduces, compares, and discusses knowledge engineering
tools that allow text-based editing of \textsc{pddl} files to set the
stage for \mypddl (Table~\ref{tab:comparison}).

\pddlstudio \cite{chomuttool,plch2012inspect} is an IDE
(\emph{integrated development environment}) for creating and managing
\textsc{pddl} projects, that is, a collection of \textsc{pddl}
files. Its main features are syntax highlighting, error detection,
context sensitive code completion, code folding, project management,
and planner integration. Many of these features are based on a parser,
which continuously analyzes the code and divides it into syntactic
elements. These elements and the way in which they relate to each
other can then be identified. The syntax highlighter is a tool that
colors constructs according to their syntactical meaning within the
code. In the case of \pddlstudio, it colors names, variables, errors,
keywords, predicates, types, and brackets each in a different
customizable color.  \pddlstudio's error detection can recognize both
syntactic (missing keywords, parentheses, etc.) and semantic (wrong
type of predicate parameters, misspelled predicates, etc.)
errors. This means that \pddlstudio can detect errors due to a
mismatch between domain and problem file in real time. The code
completion feature offers recommendations for standard standard
\textsc{pddl} constructs as well as for previously used terms. Code
folding allows the knowledge engineer to hide currently not needed
code blocks. In this case only the first line of the block is
displayed. All these above mentioned features of \pddlstudio utilize
the parser. To keep track of all files, to display them in a tree
structure, and to save them upon compilation, a project manager is
integrated. Furthermore, this feature is necessary to detect interfile
errors and to use the code completion functionality. Lastly, a
command-line interface allows the integration of planners in order to
run and compare different planning software.

Unlike \pddlstudio, which provides a text based editor for \pddl, the
\itsimple \cite{vaquero2005itsimple} editor has, as its main feature,
a graphical approach that allows designing planning tasks in an
object-oriented approach using \uml (\emph{Unified Modeling
  Language}). In the process leading up to \itsimple,
\textsc{uml.p}---\textsc{uml} in a Planning Approach---was proposed, a
\textsc{uml} variant specifically designed for modeling planning
domains and problems \cite{vaquero2006use}.

The main purpose of \itsimple is supporting knowledge
engineers in the initial stages of the design phase by making tools
available that help transform the informality of the real world to
formal specifications of domain models. The professed aim of the
project is to provide a means to a \enquote{disciplined process of
  elicitation, organization and analysis of requirements}
\cite{vaquero2005itsimple}. However, subsequent design stages are also
supported. Once domain and problem models have been created,
\textsc{pddl} representations can be generated from the \textsc{uml.p}
diagrams, edited, and then used as input to a number of different
integrated planning systems. Compared to \pddlstudio, \itsimple has a
different approach to planner integration: domain and problem can be
fed to the planner by pressing a button, whereas in \pddlstudio, the
user has to know and input commands in a command-line interface.

With \itsimple, it is possible to directly input the domains and
problems into a planner and to inspect the output from the planning
system using the built-in plan analysis. This consists of a plan
visualization, which shows the interaction between the plan and the
domain by highlighting every change caused by an action. \itsimple's
modeling workflow is unidirectional as changes in the \pddl domain do
not affect the \uml model and \uml models have to be modeled manually,
meaning that they cannot by generated using \pddl. Starting in version
4.0 \cite{vaquero2012itsimple4} \itsimple expanded its features to
allow the creation of \pddl projects from scratch (i.e. without the
\uml to \pddl translation process). Thus far, the \pddl editing
features are basic. A minimal syntax highlighting feature recognizes
\pddl keywords, variables, and comments. Furthermore, \itsimple
provides templates for \pddl constructs, such as requirement
specifications, predicates, actions, initial, and goal definitions.

Starting in version 4.0, \itsimple expanded its features to allow the
creation of \pddl projects from scratch, that is, without the \uml to
\pddl translation process \cite{vaquero2012itsimple4}. So far, a basic
syntax highlighting feature recognizes \pddl keywords, variables, and
comments. \itsimple also provides templates for \pddl constructs, such
as requirement specifications, predicates, actions, initial state, and
goal definitions.

\pddlmode\footnote{\url{http://rakaposhi.eas.asu.edu/planning-
    list-mailarchive/msg00085.html}} for Emacs builds on the
sophisticated features of the widely used Emacs editor and uses its
extensibility and customizability. \pddlmode provides syntax
highlighting by way of basic pattern matching of keywords, variables,
and comments. Additional features are automatic indentation and code
completion as well as bracket matching. Code snippets for the creation
of domains, problems, and actions are also available. Finally,
\textsc{pddl}-mode keeps track of action and problem declarations by
adding them to a menu and thus intending to allow for easy and fast
code navigation.

@@

The online tool \epd allows for editing \pddl files in a web
browser. Its feature comprise syntax highlighting, code folding,
\pddl-specific auto-completion, and multi-tab support. The editor is
part of the Planning.Domains initiative which aims at providing three
pillars to the planning community: (1) an API to access existing \pddl
domains and problems; (2) a planner-in-the-cloud service; and (3) an
online PDDL editor. The online editor is also connected to the
planner-in-the-cloud.

\epd's syntax highlighting features are basic; so far, it does not
support context-aware syntax highlighting: for example,
editor.planning.domains highlights \pddl key words even if they are
misplaced. In contrast, \mypddl is aware of the scope of a construct
and highlights key words accordingly.

The PDDL plugin
vscode-\pddl\footnote{\url{https://marketplace.visualstudio.com/items?itemName=jan-dolejsi.pddl}}
for the editor Visual Studio Code offers a wide range of editing
functions, such as syntax highlighting, code completion, and code
snippets. It offers a mature planner integration and plan
visualization. Thanks to a \pddl parser integration, it is possible to
detect semantic errors immediately when they are made.

\subsection{Critical Review}
\label{sec-3-4}

All the above-mentioned tools provide environments for the creation of
\pddl code. Their advantages and disadvantages are reviewed in this
section. At the end of each discussed feature, the approach that was
used in \mypddl is introduced.

\pddlstudio, \itsimple, and \epd for the most part do not build on
existing editors and therefore cannot fall back on refined
implementations of features, such as selection of tab size, defining
custom key shortcuts, customizing the general look and feel, and
bracket matching. In contrast, vscode-\pddl is integrated into the
mature editor Visual Studio Code and can be used in combination with
other plugins. To have both basic editor features and a high
customizability, it was decided to use an existing, extensible text
editor to integrate \mypddl into.

The tools can also be compared in terms of their syntax highlighting
capabilities. In \pddlmode for Emacs (up to \pddl\,2.2) and \epd (up
to \pddl\,3.1), and vscode-\pddl (up to \pddl\,3.1) keywords,
variables, and comments are highlighted. However, this is only done
via pattern matching without controlling for context. This means that
wherever the respective terms appear within the code they will get
highlighted, regardless of the syntactical correctness. Therefore, it
is useful when the knowledge engineer is familiar with \pddl syntax
but can also be misleading if this is not the case. Different colors
can be chosen by customizing Emacs and Visual Studio Code. \epd
provides two fixed color schemes. \itsimple's syntax highlighting for
\pddl 3.1 is, except for the \pddl version difference, equally as
extensive as that of \pddlmode for Emacs but does not allow for any
customization.  \pddlstudio has advanced syntax highlighting that
distinguishes all different \pddl 1.2 constructs depending on the
context and allows knowledge engineers to choose their preferred
highlighting colors. One of the primary objectives of \mypddl is to
help users in keeping track of their \pddl programs. As a means to
this end, it was decided to also implement sophisticated,
context-dependent syntax highlighting.

Another feature that can be useful for fast programming is the ability
to insert larger code skeletons or snippets. This allows the knowledge
engineer to focus on the specific domain and problem characteristics
instead of having to worry about the \pddl formalities. \pddlstudio
does not support the insertion of code snippets. \itsimple features
some code templates for predicates, derived predicates, functions,
actions, constraints, types, comments, requirements, objects, and
metrics. However, the templates are neither customizable nor
extensible. \pddlmode for Emacs provides three larger skeletons: one
for domains, one for problems, and one for actions. Further skeletons
could be added. Both \epd and vscode-\pddl provide many code
snippets. \mypddl aims to combine the best of these latter tools and
support customizable and extensible snippets for domains, problems,
types, predicates, functions, actions, and durative actions.

\pddlstudio, \pddlmode for Emacs, and \epd do not provide
visualization options. \itsimple, on the other hand, is based entirely
on visually modeling domains and problems. Therefore, since the first
version, the focus has mainly been on exporting from \textsc{uml.p} to
\pddl. \mypddl is to reverse this design approach and enable type
diagram visualization of some parts of the \pddl code. vscode-\pddl
does not provide domain visualizationa but is able to visualize a
found plan.

Searching for errors can be one of the most time consuming parts of
the design process. Hence, any tool that is able to help detect errors
faster is of great value to the knowledge engineer. While \pddlmode
for Emacs, \itsimple, and \epd facilitate error detection only by
basic syntax highlighting, both \pddlstudio and vscode\-pddl are able
to detect errors via a \pddl parser. In \mypddl, syntactic errors are
not be highlighted by the syntax highlighting feature, while all
correct \pddl code is to get highlighted.

A major drawback of \pddlstudio and \pddlmode for Emacs especially is
that they are not updated regularly to support the most recent \pddl
versions. \pddlstudio's parser is only able to parse \pddl 1.2
\cite{chomuttool}, while the latest \pddl version is 3.1. \pddl has
significantly evolved since \pddl 1.2 and was extended in \pddl 2.1 to
include \emph{durative actions} to model time dependent behaviors,
\emph{numeric fluents} to model non-binary changes of the world state,
and \emph{plan-metrics} to customize the evaluation of plans
\cite{fox2003pddl2}. \pddlmode for Emacs is only compatible with \pddl
versions up to 2.2, which introduced \emph{derived predicates} and
\emph{timed initial predicates} \cite{edelkamp2004pddl2} but does not
recognize later features like \emph{object-fluents}. It follows that
the range of functions specified in the domain file cannot include
object-types in addition to numbers. \itsimple, \epd, vscode-\pddl,
and \mypddl support the latest \pddl version.

An important features of any software is the possibility of extending
and customizing it \cite{hui2003requirements}. \pddlstudio falls short
of satisfying this requirement as the customization options are
limited to the choice of font style and color of highlighted \pddl
expressions. Furthermore, \pddlstudio is written as standalone
program, meaning that there are no \pddl independent extensions. The
same holds true for \itsimple. Since both Emacs and VS Code are
established editors, \pddlmode and vscode-\pddl are highly and easily
customizable and extensible. This is the other major reason why it was
decided that \mypddl should be integrated into a existing, extensible,
and customizable text editor. These requirements are intended to be
met by Sublime Text, a text editor that offers such features as
customizable key bindings, display of line numbers, and multi-line
selection.

All in all, \mypddl must be understood as complementary to the other
existing knowledge engineering tools. \mypddl is distributed as a
package for Sublime Text and provides context-aware syntax
highlighting, code snippets, syntactic error detection, and type
diagram visualization. Additionally, it allows for the automation of
modeling tasks due to an interface with Clojure that supports the
conversion of \textsc{pddl} code into Clojure code and vice
versa. Therefore, \mypddl is intended to support both the initial
design process of creating domains with code snippets, syntax
highlighting and the Clojure interface and the later step of checking
the validity of existing domains and problems with the type diagram
generator. Lastly, the visualization capabilities of \mypddl are meant
to facilitate collaboration among knowledge engineers.
